GEE models

### 2017-09-25 doc/submission/reviewer-response-2.md

---
output: fost::default_manuscript
---

# Response to reviewers

## Referee: 1

- *"This paper presents an investigation of the relationship between NEFAs measured
at baseline and subsequent development of markers of insulin sensitivity. The
analysis is very thorough and the conclusions, in particular the interpretation
of results with biological considerations is excellent. I am not an expert on
this subject area and although I am a statistician, GEE is not an approach that
I am familiar with, so my comments are from the perspective of a
technically-minded but non-expert reader. Therefore some of my comments may not
be wholly relevant and are given more as possible considerations; or areas that
may require clarification for other non-experts."*

- *"A key question for me is the choice of using the GEE approach, which is of
course an established and appropriate method for many research questions, but I
think could be better justified here. The main advantages I am aware of are:
Richer correlation structures, easily handles missing data, robust standard
errors."*

- *"Here, the number of follow up visits is not large and the amount of missing data
is relatively small. My first thought when approaching the analysis would
therefore be to consider a standard linear regression analysis with
robust/clustered standard errors to incorporate repeated measures, and potential
multiple imputation of missing values and a comparison with complete case
analysis. So it would be nice to briefly spell out the things that GEE offers
here over other approaches to make this clear (in the ESM if need be so as not
to add to the word count). Currently the main statement is the ease of handling
missing data – which is handy but perhaps not essential – and some comments on
the working correlation matrix, which are not terribly informative. So some
statements about the improvement in estimates under a specific correlation
structure that is not easily handled otherwise, for instance, might be helpful.
Of note is that in the ESM it is stated that "other … matrices had similar fit"
– it may be worth noting that the model fit may be similar, but estimates may be
more efficient under a particular correlation structure."*

- *"Another thing that might be commented on briefly is the different nature of
estimates from GEE models, which are population averaged and have a slightly
different interpretation. This difference might be seen as a disadvantage to GEE
modelling – or do the authors feel this is more appropriate for this particular
question? This is something I do not have a thorough understanding of, but I do
note that the authors comment in the methods "…the interpretation of the GEE
results is as an expected percent difference in the outcome variable for every
standard deviation (SD) increase in the predictor variable given the covariates
are held constant" does not appear to be consistent with what I have read in
terms of keeping other factors constant, e.g.: "However, because the GEE
approach does not contain explicit terms for the between-cluster variation, the
resulting parameter estimates for the contrast of interest do not have the usual
"keeping other factors constant" interpretation." (Hanley et al, Am J Epidemiol
2003;157:364–375)."*

- *"Another area that is unclear is the exact model specification. The unadjusted
results I can assume are the outcome variables regressed via GEE against each
NEFA measure. Then adjusted results include age, gender, WC and so on. Now, for
the proportion of NEFA results, the approach seems to be that adjusted models
were fitted with each NEFA as a predictor in turn while adjusting for the
anthropomorphic and other covariates, as a PLS approach was used to examine all
the NEFA components together as proportions. Was this approach the same for the
analysis of NEFA as a concentration, or in this analysis were all NEFA
components included in a single multivariable model? I think that this should be
spelled out: if each NEFA component was considered in turn, and the variables
adjusted for, for the concentration/proportion analysis. At present the
assumption might be that multivariable models include all of the NEFA components
in one model."*

- *"If so, I do not fully understand why there is not a model that includes the
joint effect of all NEFA components together. Could this not be done for the
concentration analysis? Also, it could perhaps be done for the proportion too,
if one of the components is set as the baseline – similar to how categorical
variables are handled in regression analyses, but allowing percentages for each
indicator rather than the usual 0/1 dichotomy, vs. a "baseline" component. Such
an analysis could also adjust for total NEFA, with the interpretation that
proportion of each NEFA component is assessed while accounting for other NEFA
components, and also the total NEFA. This is also something I struggled to
understand a little in general – why should *proportions* of NEFA components be
important? There may be a relatively high proportion of a particular NEFA but if
the patient has a generally low overall NEFA, is there any reason to think a
relatively high proportion of something will increase insulin sensitivity? These
ideas may be wholly implausible of course and I may have missed the point – as I
said my background knowledge of the biology is not ideal."*

- *"I do also wonder somewhat whether there may be over-adjustment with some of the
covariates included, and agree with another reviewer who pointed out the
importance of the unadjusted results. A potential issue is confounders that may
be on the causal pathway, which is a judgement that must be based on expert
knowledge. The authors note for instance that adjusting for WC improved the fit
of the model, and was indicated for inclusion by the DAG. Neither in my mind is
reason alone for its inclusion, and there is no guarantee on this basis that
adjustment for this is necessarily appropriate – this must be based on
subject-specific knowledge of whether adjustment is appropriate to determine the
independent effect of the variable of interest."*

- *"Which brings me to the DAGs. "Directed acyclic graphs" just means a collection
of boxes representing different bits of information with arrows between them
that do not form loops; DAGs are used in Bayesian analysis for instance to
represent how data and parameters in a model are linked. A DAG therefore is a
representation for how information in linked, not a process to in itself to
determine appropriate model building or make a "recommendation". I think that if
the approach employed is to be used as justification of the model building
process, it needs to be better described: presumably there is some algorithm
being used that determines interrelationships of variables in the data, of which
I would like to see at least a brief outline – a link to a website is not really
sufficient here. Further, the graphs provided give no indication of what the
different arrows and colours are."*

- *"Last of all, the analysis excluded diabetes cases, which is justified. However,
could incident cases be included with follow up time included to the last visit
prior to developing diabetes? This may not be so straightforward in practice of
course. It’s not really clear if any of any pre-incident follow up time is
included at present, or if all data on incident cases is discarded "incident
diabetes cases were excluded from further GEE analysis" – does the "further" bit
mean post-incidence here?"*

- *"MINOR COMMENTS:"*
    - *"Abstract and elsewhere: OGTT not spelled out, even if this is obvious to
    most readers."*
    - *"List of abbreviations: if this is to be an exhaustive list many are not
    given, OGTT, HOMA, NEFA"*

    - *"QIC - presumably this is quite like the AIC and other information criteria -
    maybe mention this, and how the calculation differs? Is the QIC specifically for
    GEE models? Also in the ESM: "This identifies a specific model from a number of
    models that is the 'best' fit and least complex of all the models" - it can’t be
    the best fit AND least complex, as mentioned in the previous sentence it’s a
    trade-off - so maybe omit this sentence."*

    - *"ESM - collider variables. As mentioned in main comments this is all a
    bit vague - "arrows appearing to collide" and so on doesn’t really convey
    what is going on here. This is a technical description so please be a little
    more precise!"*

## Referee: 2

- *"In this manuscript, the authors investigated the link between total and several
fractions of serum NEFA and the clinical indexes of insulin secretion and
sensitivity by using GEE model in a diabetes risk group over 6-years
longitudinal follow-up."* ... *"Overall, this study is interesting and novel to
show the long-term exposures to some specific NEFA fractions are negatively
associated with mainly beta-cell function in human, which convincingly confirms
and advances the previous basic and clinical studies in this field."*

1. *"Serum insulin and/or NEFA levels could be affected by several factors: I
would suggest excluding the participants who have endocrine, hepatic or renal
disorders. Also, I suggest mentioning the detail of medications, which could
affect insulin/glucose/NEFA measurement; page 24, line 7, "Medication use did
not influence NEFA composition, the outcomes, or the GEE models (data not
shown)" is not clear enough."*

2. *"Glucose clamp test is the standard to evaluate insulin sensitivity instead
of the indexes used in this study. I know it is very hard to perform it in a
large number of participants, but it may be one of the reasons why the effect of
NEFAs on insulin sensitivity was limited. I would suggest discussing the
limitation of these indexes."*

3. *"Most of the indexes in this study are based on glucose and insulin levels.
I think the reader would know the data of baseline and follow-up glucose and
insulin levels as well as indexes (in Table 1)."*

4. *"I understand several NEFA fractions are statistically related to indexes of
beta-cell function; however, does each individual show the same trend? For
example, the graph which shows NEFA concentrations and the individual decline
ratio of beta-cell function over 6-years (value of slope, if compatible) may
convince the reader more, at least for the representative NEFA fractions."*

5. *"Referring previous longitudinal studies showing the link between NEFA level
and beta cell deterioration, even which were investigated in the diabetic
patients, could confirm the results of this study."*

## Referee: 3

- *"The authors addressed all major concerns related to the manuscript. I feel
that the paper improved by the changes and became more focused by omitting the
PLS-DA analysis. By adding a detailed explanation to the discussion on the
potential importance of potential confounders and mediators, the interpretation
of the findings became much clearer."*

- *"Some minor comments regarding the wording of the conclusions and missing
information on the DAGs remain but I only have the following comments:"*

- *"Minor comments"*
    1. *"Introduction: "Therefore, our objective was to examine the association of
    serum NEFA composition on changes over time in insulin sensitivity and beta-cell
    function in a longitudinal cohort. We hypothesized that higher palmitic acid and
    lower PUFA such as eicosapentaenoic acid in the NEFA fraction would associate
    with declining insulin sensitivity and beta-cell function over 6 years." I
    would prefer to use the word differences instead of changes, as the main
    focus of the paper is not on the time interaction but on the main effects."*

    2. *"Conclusions: "In a Canadian population of adults who were at-risk for
    diabetes, we found that higher total NEFA concentrations independently predicted
    lower beta-cell function after 6 years." I would prefer to use over 6 years
    instead of after 6 years, as the difference was present from the measurement
    point to the end of follow-up."*

    3. *"Please add the number of observations for each clinic visit in Table 1."*

    4. *"Please add the number of incident and prevalent diabetes cases to the
    respective exclusions in ESM Figure 1. For example: n=736 had baseline data
    collected, n=XX diabetes cases were excluded; n=619 had 3-year data, n=XX
    incident diabetes cases were excluded..."*

    5. *"Please add the colour codes for DAGs (i.e.: exposure, outcome, ancestor
    of exposure, ancestor of outcome, ancestor of both, causal path, biasing
    path)."*

## Referee: 4

- *"This is an interesting analysis on NEFA ,insulin sensitivity and beta cell
function."* ... *"My major concern is that with the current approach the
complexity of metabolic regulation may be lost to some extent. This makes it
hard to interpret these findings with respect to biological significance:"*

    - *"With respect to peripheral insulin sensitivity (referred to in the
    discussion) the effects of saturated fatty acids may strongly depend on
    metabolic phenotype, ie more pronounced in insulin resistant subjects, was
    initial metabolic phenotype taken into account in the present analysis?"*

    - *"it is a pity that the analysis of fatty acids composition in
    phospholipids and serum cholesteryl esters have been published in a separate
    paper, altogether this would add very much to the total interpretation."*

    - *"some discussion on what determines plasma fatty acid profile may be
    useful, beside spillover also (selective) release by endogenous lipolysis.
    To what extent does it reflect dietary composition, was habitual fat intake
    taken into account?"*

- *"Altogether, this paper would gain strength by addressing the limitations in
interpretation in more detail."*

## Associate Editor

- *"The reviewers provide numerous minor and major comments, for example regarding
motivation of the statistical approach, which require revision or careful
response."*

### 2017-09-29 doc/manuscript.Rmd
changes to differences

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

@@

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
epidemiological studies that only looked at total NEFA and not individual fatty 
acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted
natural history of diabetes and the growing appreciation of the divergent
effects of individual fatty acids. Therefore, our objective was to examine the
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
NEFA fraction would associate with declining insulin sensitivity and beta-cell
function over 6 years.

>>>>

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

@@

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
epidemiological studies that only looked at total NEFA and not individual fatty 
acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted
natural history of diabetes and the growing appreciation of the divergent
effects of individual fatty acids. Therefore, our objective was to examine the
association of serum NEFA composition on differences over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
NEFA fraction would associate with declining insulin sensitivity and beta-cell
function over 6 years.
